Modeling Human Interactions in Collaborative Interactive Evolutionary Computation

### 2017-02-02 2017-GECCO-Modeling-User-Interactions-in-IEC/paper-body.tex
Results

% IEAs
Interactive evolutionary computation (IEC) systems are, in general, evolutionary methods
whose fitness evaluations are performed by humans through an
interactive system \cite{eiben2015interactive}. They are usually
applied in problems where the fitness function is not known  or simply
does not exist, and the result
of optimization should fit certain human and possibly aesthetic
need. That is why their use case includes the evolution of objects
with subjective characteristics such visual appeal or 
attractiveness \cite{biomorphs} as well as others where human behavior is 
considered, for instance to optimize teamwork \cite{kosorukoff2002evolutionary}
or creativity cite[cook].
% Human Based Computation
These IEC systems are an interesting venue of research, since they have demonstrated their ability for effectively 
producing art and design \cite{Bentley:1999:intro,Sims:1991,todd:1992,evoeco},
as well as other artifacts in many other domains \cite{ie1}. In these cases when human interaction is responsible of other 
operations of the evolutionary process some authors classify these IEC methods 
as human-based evolutionary computation cite[kosorukoff] or as human-based computation cite[].

% C-IEAs
One of the issues with these systems is that their performance effectively depends on the number of users
they are able to include; and in order to reach more human users,
Interactive Evolutionary Algorithms (IEAs) are 
some times developed as web applications depending on visitors to the web to interactively help
with the search, using both anonymous and registered users. Some systems 
employ a collaborative technique, where several users assign a quality assessment 
to a single solution and then an aggregated fitness value is calculated 
\cite{picbreeder,seyama2016development,wagy2014collective}.
% Problems
However, the necessary intervention of humans in these systems leads
to them having several inherent problems arising from the very nature of 
the algorithms, namely, human evaluations are slow and expensive, there is a
human fatigue caused by the interaction \cite{ie1}, and
the boredom arising when users evaluate a large number of phenotypes 
many of which are not interesting or are very similar to each other. 
The number of evaluations that IEC can receive is limited by the number of users
collaborating and the amount of interaction of each. Having a volunteer system 
can lower the requirements of anyone participating in
the experiment thus increasing the {\em performance}, in terms of {\em
human brain cycles}, of the whole system.
% Volunteers

Using a volunteer based system raises other issues \cite{sarmenta2001volunteer,web:BOINC}, 
such as the volunteer\'s lack of accountability,
and the need to build trust between participants and application
providers. Issues for the provider of the system or organizer  are also the difficulty of establishing 
the amount of time and resources
a volunteer is willing to spend on the system, and how they decide if they
participate or not \cite{JJ:2016}.

% Human Centered
In order to increase volunteer participation and to tackle some of the issues mentioned above,  
we proposed a framework following a human centered design \cite{greenhouse2012human},
giving extensive attention to volunteers, not only because their
explicit evaluation is essential, but also because the context of the 
interaction affects the system as a whole. In collaborative systems, users interact 
with each other, and are aware of the actions ref[]. Users also consider their previous evaluations
and past experiences ref[].

%Please help with this paragraph, is evidence needed?
% It's the tuning of the algorithm which is data driven, not the
% algorithm itself, right? - JJ
%  Tuning yes - Mario
% And evidence for what? - jj
%  That most EAs use only population data - Mario
In evolutionary computation algorithms are data driven since it is important to keep data about the 
population of each generation, in order to generate the next or even change certain parameters
or analyze the evolutionary process. 
% Human Centered Importance
The framework models the context of interaction along with the population and
this knowledge is available to be an integral part of the evolutionary computation.
%This could be generalized?
For example, in a C-IEC application, fitness assignment depend on the
actions of a social network of users. 
A stream of actions is triggered when they tag, share, rate, store or even delete a phenotype. 
The selection of parents could also depend also on the previous actions, leveraging information 
such as the fact that they have the same tag, or are shared by
similar users. That knowledge could also be applied in choosing which phenotypes present to a certain
users, in order to give them those they will find more interesting or useful. 
% IoT
The context of interaction, can also include data retrieved from sensors and other IoT devices.
% Engagement
Data available from the interaction is also used to increase the performance of the system by applying 
gamification techniques. Gamification is defined by Deterding et al. \cite{deterding2011game} as
\begin{quote}
  the use of game design elements in non-game contexts.
\end{quote}  
The gamification element employed in this case study has been a rewarding mechanism  
\cite{dubois2013understanding}. In general rewards  consist of a reputation system with score points, 
levels and leader boards. Points are awarded to users in response of 
the accomplishment of certain activities that need to be encouraged. Levels depend
on the score and certain features of the game are only available to gamers when 
they reach a giving level. % Say what we are doing in this case - JJ

Three case studies are presented as a proof of concept, providing both
conceptual and implementation details of the framework. 
The experiments were implemented using the EvoSpace-Interactive (ES-I) framework \cite{garcia2013evospace}, 
and where employed in different contexts of interaction. 
Additional details of each case study application is presented elsewhere refs[][][]. 
Giving this paper the objective of presenting the underlying conceptual framework.

@@

\section{Related Work}
\label{sec:related}
% C-IEAS
Given the fatigue fitness evaluations produce in human users when applying IEAs, some authors have tried to mitigate
the problem by allowing the algorithm to collaborate with the user, so that sometimes users perform the evaluation, 
but also specific measures are included into the algorithm to perform the evaluation of some features automatically.  
For instance, Reis et al. added some terrain measures (such as accessibility and edge length) to a standard 
IEC in \cite{DBLP:journals/soco/FradeVC12} .  This way the algorithm was capable of providing terrains that would
otherwise have need users evaluation for these specific features.

% Volunteer Based
% HBC
% Engagement Techniques
% Models

@@

\subsection{Graph}
  In order to deal with large datasets of connected data found in these systems,
  graph databases \cite{angles2012comparison} have been proposed as an alternative to relational databases 
  which have performance limitations when dealing with highly connected data \cite{holzschuher2013performance}.
  A graph is proposed for modeling the social network of users, their interaction with 
  candidate solutions, and the relationships between them in the population.
  The graph database system used in the implementation is Neo4J as it is
  a scalable solution \cite{miller2013graph,holzschuher2013performance}, and well supported and documented
  in PaaS infrastructures like Heroku. The Cypher query language it used to retrieve views from the graph.
  An exaplmple query is shown in figure \ref{fig:cypher} where the relations between users and solutions 
  are presented.

@@

\subsection{PostgreSQL}
  The PostgreSQL relational database system is also employed because user sessions and authentication,
  as well as dynamic web pages are handled directly by the Django web
  framework \cite{garcia2013evospace}. % these implementations details
                                % could go if needed.

@@

\subsection{Evolution}

\section{Case Studies}
\label{sec:experiments}

@@

\subsection{EvoDraw}
As a case study, a IEA was developed using the 
ES-I platform \cite{garcia2013evospace}.
A brief description of the application is presented next, focusing
on the data elements that where ported to the graph model.

\subsection{Fitness Assignment}
\label{sec:assignment}
% changed to this, because I understand it does not always work that
% way - JJ
The ES-I platform has been programmed to employ a collaborative methodology for performing fitness
 assignment. Several registered users assign a quality assessment to a single
phenotype and then an aggregated fitness value is calculated,
depending on the rating (from 1 to 5 stars) by each particular user,  % how do you define
                                % "taste"? - JJ
                                % Using actions
resulting in a many-to-many relationship between users and
phenotypes. 
 %scored by them? - JJ 
Other components or users of the systems can 
query this user-phenotype relation to extract relevant
knowledge about the process and the population, for instance showing the
most popular phenotype, or the the user with more participation 
\cite{picbreeder}.
In order to do this, metadata about phenotypes 
must be permanently stored, even
if they are no longer in the active population.

\subsection{Collaboration}
\label{sec:col}
Users need to authenticate themselves to the system using their
Facebook account. In that sense, some users might not be interested
either because they do not want to give that information or simply because
they do not use that particular social network. Even as we might lose
some users that way, the additional information we obtain for scoring
phenotypes more than compensates for that. % Added this, check it
                                % out.  
                                % Ok! - Mario (and knowing how much more is future work)
After entering the web application by using their Facebook account,
users can collaborate with their Facebook friends, 
sharing those phenotypes they like, or by taking phenotypes
from their friend's collections by using the web interface depicted 
in Figure \ref{fig:web}.
At the top left corner a list of Facebook friends is presented
to encourage users to interact with the system. In the central 
\emph{ Wall } area, a phenotype sampled from the population that is
being evolved via the evolutionary algorithm 
is shown to the user.
Here, the user can interact with the system in two ways.
First, he can assign a rating % Why don't you use other
                                % thing instead of quality? We are not
                                % talking about quality, but about
                                % tastes or similar - JJ
                                % is rating Ok?
% Rating is probably OK. 
 to the phenotype using
a five star rating system or,
additionally, a user can choose to add an image to one of their \emph{Collections}.
A collection is a special folder that stores those phenotypes a user likes and wishes
to save. After the user finishes interacting with the phenotype
on the Wall, he can choose to retrieve a new one from the population.
At the left hand side of \ref{fig:web}, the web page shows the \emph{Collections} section.
The user can create several collections, to group and organize his favorite 
artifacts. Moreover, users can browse the content of each collection and from
there share images through their social network.
When a user hovers the cursor over a phenotypes a pop-up pane shows how many users have
liked the phenotype. The pane also includes a link to the phenotype's 
details, the parents, genetic operators that created it, and genealogy
information. This makes the assignment of fitness through the rating
system a {\em social} activity, pursuing the objective of this work,
which is to increase engagement of the users.
\subsubsection{Graph Model}

\subsubsection{Gamification}

\subsubsection{Setup}
\subsubsection{Results}

@@

\subsection{EvoDraw-Kinect}
\subsection{XYZ-Unplugged}

@@

>>>>

% IEAs
Interactive evolutionary computation (IEC) systems are, in general, evolutionary methods
whose fitness evaluations are performed by humans through an interactive system \cite{eiben2015interactive}.
They are usually applied in problems where the fitness function is not known  or simply
does not exist, and the result of optimization should fit certain human need. That is why their use 
case includes the evolution of objects with subjective characteristics such visual appeal or 
attractiveness \cite{biomorphs} as well as others where human behavior is 
considered, for instance to optimize teamwork \cite{kosorukoff2002evolutionary}
or creativity cite[cook].
% Human Based Computation
These IEC systems are an interesting venue of research, since they have demonstrated their ability for effectively 
producing art and design \cite{Bentley:1999:intro,Sims:1991,todd:1992,evoeco},
as well as other artifacts in many other domains \cite{ie1}. In these cases when 
human interaction is responsible of other 
operations of the evolutionary process some authors classify these IEC methods 
as human-based evolutionary computation cite[kosorukoff] or as human-based computation cite[].
% Problems
However, the necessary intervention of humans, brings certain challenges to designers of IEC methods  
namely, human evaluations are slow and expensive, there is a
human fatigue caused by the interaction \cite{ie1}, and
there is also the boredom arising when users evaluate a large number of phenotypes 
many of which are not interesting or are very similar to each other.
% C-IEAs
Moreover, the performance of theses systems effectively depends on the number of users
they are able to include; and in order to reach more human users,
Interactive Evolutionary Algorithms (IEAs) are 
some times developed as web applications depending on visitors to the web to interactively help
with the search, using both anonymous and registered users. Some systems 
employ a collaborative technique, where several users assign a quality assessment 
to a single solution and then an aggregated fitness value is calculated 
\cite{picbreeder,seyama2016development,wagy2014collective}.
%The number of evaluations that IEC can receive is limited by the number of users
%collaborating and the amount of interaction of each. 
Having a volunteer system  can lower the requirements of anyone participating in
the experiment thus increasing the {\em performance}, in terms of {\em
human brain cycles}, of the whole system.
% Volunteers 
Using a volunteer based system raises other issues \cite{sarmenta2001volunteer,web:BOINC}, 
such as the volunteer\'s lack of accountability,
and the need to build trust between participants and application
providers. Issues for the provider of the system or organizer  are also the difficulty of establishing 
the amount of time and resources
a volunteer is willing to spend on the system, and how they decide if they
participate or not \cite{JJ:2016}.

% Human Centered
In order to increase volunteer participation and to tackle some of the issues mentioned above,  
we proposed a software framework following a human centered design \cite{greenhouse2012human},
giving extensive attention to volunteers, not only because their
explicit evaluation is essential, but also because the context of the 
interaction affects the system as a whole.
%Evolutionary computation algorithms are data driven since it is important to keep data about the 
%population of each generation, in order to generate the next or even change certain parameters
%or analyze the evolutionary process. 
% Human Centered Importance
%The framework models the context of interaction along with the population and
%this knowledge is available to be an integral part of the evolutionary computation.
%This could be generalized?
For example, in a C-IEC application, fitness assignment depend on the
actions of a social network of users.  Actions are triggered when 
they tag, share, rate, store or even delete a phenotype. 
The selection of parents could depend on the previous actions, leveraging information 
such as the fact that they have the same tag, or where shared by
similar users.
% IoT
%The context of interaction, can also include data retrieved from sensors and other IoT devices.
% Engagement
Data available from the interaction is also used to increase the performance of the system by applying 
gamification techniques. Gamification is defined by Deterding et al. \cite{deterding2011game} as
\begin{quote}
  the use of game design elements in non-game contexts.
\end{quote}  
The gamification element employed in this case study has been a rewarding mechanism  
\cite{dubois2013understanding}. In general rewards  consist of a reputation system with score points, 
levels and leader boards. Points are awarded to users in response of 
the accomplishment of certain activities that need to be encouraged. Levels depend
on the score and certain features of the game are only available to gamers when 
they reach a giving level. % Say what we are doing in this case - JJ

Three case studies are presented as a proof of concept, providing both
conceptual and implementation details of the framework.

@@

\section{Related Work}
\label{sec:related}

An early example of web based collaboration is the work by 
Langdon \cite{langdon:2004}, which evolved fractal representations of virtual creatures. 
Similarly, Secretan et al. \cite{picbreeder} and Clune and Lipson \cite{forms} use web-based IEAs 
to evolve artistic artifacts using a generative encoding, compositional pattern producing networks.
Another example was the Galapagos Project \cite{sims1997interactivity},
an exhibit in the Tokyo Multimedia Museum (1997--2000) were visitors interacted with images presented in 
twelve displays by selecting those they found most aesthetically interesting by standing on
step sensors in front of them. 
% C-IEAS
Given the fatigue fitness evaluations produce in human users when applying IEAs, some authors have tried to mitigate
the problem by allowing the algorithm to collaborate with the user, so that sometimes users perform the evaluation, 
but also specific measures are included into the algorithm to perform the evaluation of some features automatically.  
For instance, Reis et al. added some terrain measures (such as accessibility and edge length) to a standard 
IEC in \cite{DBLP:journals/soco/FradeVC12} .  This way the algorithm was capable of providing terrains that would
otherwise have need users evaluation for these specific features. Seyama and Munetomo \cite{seyama2016development}
also propose the reduction of the user fatigue by using a collaborative filtering algorithm to show only the
information utilized by similar users as they collaborate with a large number of users for 
the interactive modeling of 3D glasses.

Some Collaborative IEC (C-IEC) systems promote user engagement by presenting interesting information to 
users, for instance the genetic lineage of each phenotype or the most popular or 
best rated solucion\cite{picbreeder,forms}.In a recent work by 
Wagy \& Bongard \cite{wagy2014collective} user interaction is needed for evaluating the fitness and developing
new designs of robot locomotion. Collaboration is encouraged by gamifying the system 
using the maximum distance indicator to inspire the user to try and ``beat'' previous designs. 
In any case, using gamification techniques imply that we must deal with interactive evolutionary
system as socio-technical constructs, where the social aspects are
essential to understand its dynamics. In this sense, conclusions
reached with other systems such as NodEO \cite{DBLP:conf/gecco/MereloCGCRV16} can also be applied to this
system; and applying social network techniques such as graph analysis
to their study will allow us to understand them more thoroughly.

Collaborative IEC (C-IEC) systems also need to store highly connected data, as it is common in current applications
like social networks. In order to deal with large datasets of connected data found in these systems,
graph databases \cite{angles2012comparison} have been proposed as an alternative to relational databases 
which have performance limitations when dealing with highly connected data \cite{holzschuher2013performance}.
The framework and data model for representing the interactive process of C-IEC systems is proposed next
% Volunteer Based
% HBC
% Engagement Techniques
% Models

@@

\subsection{Graph-Neo4J}
  In order to deal with large datasets of connected data found in these systems,
  graph databases \cite{angles2012comparison} have been proposed as an alternative to relational databases 
  which have performance limitations when dealing with highly connected data \cite{holzschuher2013performance}.
  A graph is proposed for modeling the social network of users, their interaction with 
  candidate solutions, and the relationships between them in the population.
  The graph database system used in the implementation is Neo4J as it is
  a scalable solution \cite{miller2013graph,holzschuher2013performance}, and well supported and documented
  in PaaS infrastructures like Heroku. The Cypher query language it used to retrieve views from the graph.
  An exaplmple query is shown in figure \ref{fig:cypher} where the relations between users and solutions 
  are presented.

@@

\subsection{Relational Objects-PostgreSQL}
  The PostgreSQL relational database system is also employed because user sessions and authentication,
  as well as dynamic web pages are handled directly by the Django web
  framework \cite{garcia2013evospace}. % these implementations details
                                % could go if needed.

@@

\subsection{Evolution}
The evolutionary algorithm is decoupled from the data model, the algorithm can be implemented as an
EvoWorker \cite{garcia2013evospace}, use an external library, or even it can be done by humans.
An additional tool was implmented for a human based evolution , where volunteers select the parents of
the next generation and upload the new individuals \cite{de2014artists}.

\section{Case Study: EvoDraw}
As a case study, a IEA was developed using the 
EvoSpace-Interactive (ES-I) platform \cite{garcia2013evospace}
A brief description of the application is presented next, focusing
on the data elements that where ported to the graph model.

@@

\label{sec:experiments}
\begin{figure}[!t]
    \centering
        \includegraphics[width=3.5in]{img/interfaceXYZ.png}
    \caption{User interface of the XYZ application.}
    \label{fig:xyz}
\end{figure}

\subsubsection{Fitness Assignment}
\label{sec:assignment}
% changed to this, because I understand it does not always work that
% way - JJ
The ES-I platform has been programmed to employ a collaborative methodology for performing fitness
 assignment. Several registered users assign a quality assessment to a single
phenotype and then an aggregated fitness value is calculated,
depending on the rating (from 1 to 5 stars) by each particular user,  % how do you define
                                % "taste"? - JJ
                                % Using actions
resulting in a many-to-many relationship between users and
phenotypes. 
 %scored by them? - JJ 
Other components or users of the systems can 
query this user-phenotype relation to extract relevant
knowledge about the process and the population, for instance showing the
most popular phenotype, or the the user with more participation 
\cite{picbreeder}.
In order to do this, metadata about phenotypes 
must be permanently stored, even
if they are no longer in the active population.

\subsubsection{Collaboration}
\label{sec:col}
Users need to authenticate themselves to the system using their
Facebook account. In that sense, some users might not be interested
either because they do not want to give that information or simply because
they do not use that particular social network. Even as we might lose
some users that way, the additional information we obtain for scoring
phenotypes more than compensates for that. 
After entering the web application by using their Facebook account,
users can collaborate with their Facebook friends, 
sharing those phenotypes they like, or by taking phenotypes
from their friend's collections by using the web interface depicted 
in Figure \ref{fig:web}.
At the top left corner a list of Facebook friends is presented
to encourage users to interact with the system. In the central 
\emph{ Wall } area, a phenotype sampled from the population that is
being evolved via the evolutionary algorithm 
is shown to the user.
Here, the user can interact with the system in two ways.
First, he can assign a rating to the phenotype using
a five star rating system or,
additionally, a user can choose to add an image to one of their \emph{Collections}.
A collection is a special folder that stores those phenotypes a user likes and wishes
to save. After the user finishes interacting with the phenotype
on the Wall, he can choose to retrieve a new one from the population.
At the left hand side of \ref{fig:web}, the web page shows the \emph{Collections} section.
The user can create several collections, to group and organize his favorite 
artifacts. Moreover, users can browse the content of each collection and from
there share images through their social network.
When a user hovers the cursor over a phenotypes a pop-up pane shows how many users have
liked the phenotype. The pane also includes a link to the phenotype's 
details, the parents, genetic operators that created it, and genealogy
information. This makes the assignment of fitness through the rating
system a {\em social} activity, pursuing the objective of this work,
which is to increase engagement of the users.

\subsection{Graph Model}

The graph model for the EvoDraw applications has the following types of
nodes: {\bf User}, {\bf Phenotype} and {\bf Collection} . The Collection node represents
a collection of drawings belonging to users. One collection can contain many phenotypes 
or be empty and a single phenotype could be shared by many collections. The interaction 
between these entities are represented by the following edges:

\begin{itemize}
\item {\bf Likes} This relation describes the interaction between a user and
a phenotype in which a rating value is assigned this is represented by 
the {\tt rate} attribute. In the current application a user is allowed to rate
a phenotype multiple times, so the date of the interactions is also of
interest.

\item {\bf Knows} The relation connects two users that know each
other in the Facebook social graph.

\item {\bf Parent} Describes what phenotype is the parent of a new
  phenotype.

\item {\bf Has} The relation describes an ownership relation between users and
those collections they own. Also a collection {\tt has}  phenotypes stored
in them.
\end{itemize}

\subsection{Gamification}
The rewarding mechanism as it is applied in EvoDraw 
consists in giving more importance to the preference of those users having a higher reputation
as given by the score points and experience levels.

The score achieved by users depends on the actions he does in the system as
each time a user does one of these actions their score is incremented by one:
\begin{itemize}
\item Start a session.
\item Rate a phenotype.
\item Create a collection.
\item Save a phenotype of the wall to a collection.
\item Save a phenotypes from a friend's collection.
\item Explore collections of other friends.
\end{itemize}

Only the ``Start a session'' and ``Explore collections of other friends'' actions 
are not reflected in the graph model since they are stored instead in the PostgreSQL relational
database system. This is because user sessions and authentication, as well as basic
web requests, are handled directly by the Django web framework \cite{garcia2013evospace}.

Two scores are used to determine the weight of a user's preference:
\begin{itemize}
\item {\em Experience}: This value depends on the score and is a value 
between 0 and 100. It is assumed for this case, that once a user
reaches a 100 actions, it has enough experience on using the application.

\item {\em Participation}: This value is simply the degree of the user node 
in the graph. This measure considers the number of users he knows,
collections and phenotypes rated.    
\end{itemize}

\section{Experimental Setup}
\label{sec:experiments}
We present in this section the results of a comparison between three versions
of the same IEC application. The goal of the experiment is to compare two 
gamification techniques to increase volunteer participation against a third 
without any technique applied. The main difference between both versions employing 
gamification is how the fitness of phenotypes
is calculated and the information presented to users.

Three versions were compared:
\begin{itemize}
\item Base (B). All users have the same weight.
\item Non Graph Gamification (G). Only experience is considered
\item Graph Gamification (GG) . Both experience and participation are considered.
\end{itemize}
When gamification was employed, all score values known where presented to users
and a ranking of users by weight was shown in a window.

The three versions of the EvoDrawing web application were deployed to the
Heroku cloud service with the same configuration: using the
Heroku Free option with a 512 MB of RAM, adding only the GrapheneDB and RedisTOGO plug-ins.
Table \ref{tab:params} shows the parameters used for the evolutionary algorithm. The graph database 
system used in the implementation was Neo4J as it was already available as a plug-in in Heroku, is well documented
and a scalable solution \cite{miller2013graph,holzschuher2013performance}.

\begin{table}
  \small
  \caption{ Parameters for experiments.  }
  \label{tab:params} 
  \centering
  \small
  \begin{tabular}{l  c   }
    \hline\noalign{\smallskip}
     Parameter & Value \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    Initial Population Size   & 80 \\ \hline
    Sample Size & 1 \\ \hline
    Step Size & 8 Samples \\ \hline
    Mutation &  \\ \hline
    Selection & Tournament \\ \hline
    Tournament Size &  6 \\ \hline
  \end{tabular}
\end{table}

% Anuncements
At the start of every experiment, a call to participation was issued
through social networks, 
particularly on Facebook and Twitter.
It is important to note that volunteers needed to log-in to Facebook and give permissions to the 
web application in order to participate, in this case there were no anonymous volunteers. This enabled the automatic access to
each participant's social network and enabled the system to store a local copy of those friends participating.
However, another experiment would be needed to see the effect this
requirement has on participation, since we are not dealing with it in
this paper.

The link used in the call for participation
was shortened Google URL, that provides metadata and analytics for the
users that click on it. 
In table \ref{tab:urls} the URL for each deployment in the Heroku platform is shown,
along with the short URL and the analytics link. In the same table a link to the GitHub 
application repository used to deploy to Heroku is also listed.    
%Isn't this anonymous? - JJ
% like in double blind? I think so, that is why no link is given in the table. - Mario

\begin{table}
  \small
  \caption{ Deployment data. URLs were {\tt
      evodrawings0(1,2,3).herokuapp.com}. For analytics URLs, add {\tt
    .info}.}
  \label{tab:urls} 
  \centering
  \small
  \begin{tabular}{l l l }
    \hline\noalign{\smallskip}
     Deployment &  Shortened URL &  Github Repository \\
    \noalign{\smallskip}\hline\noalign{\smallskip}
    B   & goo.gl/jLis4Q &   \\ \hline
    G   & goo.gl/jqjNy5 &  \\ \hline
    GG  & goo.gl/J8TCe1 &   \\ \hline
    \end{tabular}
\end{table}

Only data for the first week of deployment was considered for the experiments, and they where conducted 
between January and May of 2016.

\subsubsection{Results} 
Before launch, each deployment was first tried with a few beta testers. When applying the leader board
gamification technique for the first time a problem was found in this stage. The problem was that some 
users were cheating by giving a rating to an animation even before it was returned from the server, this was done by just
constantly clicking the mouse button. This is a common problem found in systems using leader boards because
by making the scores visible to other players they are encouraged to compete \cite{hickman2010total}. 
According to \cite{kumar2013gamification} best-practice gamification designs try to
refrain from using this element because of  problematic consequences of competition, which can result 
in negative conduct, low cooperation and collaboration, or disadvantaging certain player
demographics such as women. In order to still measure the effect of the leader board on the system, the 
problem of cheating was mitigated by disabling the rate button for a certain amount of time giving the animation
enough time to finish. But the problem persists, competitive users will try to get in to the leader board by other means.

The results of each of the three experiments in terms of participation are detailed next.

\subsection{ Base deployment}
After a week of the announcement the total number of volunteers for this experiment was 53 registered users. 
The total number of nodes generated in the graph was 595, with a total of 2220 relationships. In this
experiment a total of 500 new phenotypes where generated. The total number of evaluated
phenotypes for each volunteer is presented in figure
\ref{fig:top-ranked-participation} in blue. Users are ranked by
the number of phenotypes they rated, in the \emph{x} axis is the rank, and in the \emph{y} axis 
the number of phenotypes evaluated using a logarithmic scale.   % Ranked by... - JJ
% Done - Mario

\subsection{ Non Graph Gamification deployment}
The total number of volunteers for this experiment was very close to
the one obtained in the previous
experiment: 54 registered users. The total number of nodes generated in the graph was 648, 
with a total of 2596 relationships. These numbers are close to those obtained in the base experiment,
with a marginal increase in participation. The increased participation also helped in the creation of
more phenotypes for a total of 648. The total number of evaluated
phenotypes for each volunteer is presented in figure \ref{fig:top-ranked-participation} in red.

\subsection{ Graph Gamification deployment}
This experiment had the higher number of volunteers for for a total of 68 registered users. 
In this experiment a total of 3594 new phenotypes where generated in the experiment. 
The total number of evaluated phenotypes for each volunteer is presented in figure 
\ref{fig:top-ranked-participation} in green.

The results presented in figure \ref{fig:top-ranked-participation}, show that those users with
higher level of participation in deployments B and G, had about the same behavior, but the difference
came with users with medium level of participation, in this case participants where more engaged in deployment
G. When comparing all the experiments the deployment GG had the higher number of participation in users
of all levels.

\section{Results}
Before launch, each deployment was first tried with a few beta testers. When applying the leader board
gamification technique for the first time a problem was found in this stage. The problem was that some 
users were cheating by giving a rating to an animation even before it was returned from the server, this was done by just
constantly clicking the mouse button. This is a common problem found in systems using leader boards because
by making the scores visible to other players they are encouraged to compete \cite{hickman2010total}. 
According to \cite{kumar2013gamification} best-practice gamification designs try to
refrain from using this element because of  problematic consequences of competition, which can result 
in negative conduct, low cooperation and collaboration, or disadvantaging certain player
demographics such as women. In order to still measure the effect of the leader board on the system, the 
problem of cheating was mitigated by disabling the rate button for a certain amount of time giving the animation
enough time to finish. But the problem persists, competitive users will try to get in to the leader board by other means.

The results of each of the three experiments in terms of participation are detailed next.

\subsubsection{ Base deployment}
After a week of the announcement the total number of volunteers for this experiment was 53 registered users. 
The total number of nodes generated in the graph was 595, with a total of 2220 relationships. In this
experiment a total of 500 new phenotypes where generated. The total number of evaluated
phenotypes for each volunteer is presented in figure
\ref{fig:top-ranked-participation} in blue. Users are ranked by
the number of phenotypes they rated, in the \emph{x} axis is the rank, and in the \emph{y} axis 
the number of phenotypes evaluated using a logarithmic scale.   % Ranked by... - JJ
% Done - Mario

\subsubsection{ Non Graph Gamification deployment}
The total number of volunteers for this experiment was very close to
the one obtained in the previous
experiment: 54 registered users. The total number of nodes generated in the graph was 648, 
with a total of 2596 relationships. These numbers are close to those obtained in the base experiment,
with a marginal increase in participation. The increased participation also helped in the creation of
more phenotypes for a total of 648. The total number of evaluated
phenotypes for each volunteer is presented in figure \ref{fig:top-ranked-participation} in red.

\subsubsection{ Graph Gamification deployment}
This experiment had the higher number of volunteers for for a total of 68 registered users. 
In this experiment a total of 3594 new phenotypes where generated in the experiment. 
The total number of evaluated phenotypes for each volunteer is presented in figure 
\ref{fig:top-ranked-participation} in green.

The results presented in figure \ref{fig:top-ranked-participation}, show that those users with
higher level of participation in deployments B and G, had about the same behavior, but the difference
came with users with medium level of participation, in this case participants where more engaged in deployment
G. When comparing all the experiments the deployment GG had the higher number of participation in users
of all levels.

@@

%\subsection{EvoDraw-Kinect}
%\subsection{XYZ-Unplugged}

@@

Collaborative and volunteer based IEC systems involve the dynamic 
interaction of many entities and artifacts. Employing a multi-model approach
can benefit researchers to understand and visualize this kind of systems better.
In this paper a graph model was proposed as a complement to current
models, initially to improve the visualization of different
experiments, but in this paper we prove that this model  enables
the implementation of a gamification technique to improve engagement in
a case study. Providing a common arena where users are aware of the
activities of other users in the social vicinity, as well as using
this graph to assign fitness allows in one hand for a greater
engagement by the users and from the algorithmic point of view a more
complex way of assigning fitness and push the generation of content
forward.

One of the interesting future lines of work would be to look a bit
more closely at the behavior of users rating the system. These initial
experiments hint at a possible power law, which might indicate that
the IEC system could be self-organizing, a process that would allow it
to reach a critical state, as has been found in software repositories,
for instance \cite{Merelo2016:repomining}. The dynamics of this kind
of system are fundamentally different, and our future research will
include exploring these aspects of the system. Another line of work would be to study the negative effects of using 
gamification techniques to improve engagement, like cheating or competition.
Finally, the refinement of the proposed Human-Centered framework will need
more case studies and further multi-disciplinary research.
